---
title: "BBMRI re-quantification comparison"
author:
- name: Daniele Bizzarri
  affiliation: LUMC
date: "15/02/2022"
output: html_document
abstract: This file was done to compare the old and the requantified datasets by Nightingale Health that are available in BBMRI-nl. You will be able to 1) Analyse the missingness within the 2 datsets, 2) correctly retrieve the same names for the metabolic features in the datasets, 3) Perform some quality Control Checks, 4) Perform the spearman's correlations between the 2 datasets, 5) plot the scatterplots and 6) Compute the mortality score in the 2 datasets.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the datasets
Please insert the path to your own metabolomics datasets.
Make sure the datasets have the same IDs and that they are IDs are in the first column.
```{r load}
#Source the functions file to load the functions
source("/Rscript/functions.R")

#Load the datasets
data_old<-read.csv("Path_to_your_old_Nightingale_metabolomics_file.csv", header = TRUE, sep = ",", row.names=1, stringsAsFactors = F)
data_req<-read.csv("Path_to_your_requantified_Nightingale_metabolomics_file.csv", header = TRUE, sep = ",", row.names=1, stringsAsFactors = F)

#Checking the IDs and if the 2 datasets are in the same order
data_req<-data_req[rownames(data_old),]
identical(rownames(data_old), rownames(data_req))

dim(data_old)
dim(data_req)

#Load the metabolomics name translator
BBMRI_translator <- readRDS("/Rscript/metabolites_name_translator_BBMRI.rds")
```

## Select the metabolites
This part will make sure that the data have the same metabolites names and it will select only the ones that are of interest.
```{r select_metabolites}
#Translate the names to BBMRI standards in the old data
nam<-find_BBMRI_names(colnames(data_old), quiet=T)
nam<-nam[which(!is.na(nam$BBMRI_names)),]
data_old<-data_old[,nam$uploaded]
colnames(data_old)<-nam$BBMRI_names

#Translate the names to BBMRI standards in the requantified data
nam<-find_BBMRI_names(colnames(data_req), quiet=T)
nam<-nam[which(!is.na(nam$BBMRI_names)),]
data_req<-data_req[,nam$uploaded]
colnames(data_req)<-nam$BBMRI_names

#select only the most metabolites uncorrelated
data_old<-data_old[,MET65]
data_req<-data_req[,MET65]
```

# Missingness
To get an overview of the missingness in samples, we use print.miss.report and an heatmap indicating the missingness.
```{r miss_old}
print.miss.report(t(data_old), on_sample=T,type="missingOrZero")

#plot missingness
plot.na.heatmap(t(data_old), title="Missingness Plot old data")
```

```{r}
print.miss.report(t(data_req), on_sample=T,type="missingOrZero")

#plot missingness
plot.na.heatmap(t(data_req), title="Missingness Plot requantified data")
```

# Quality Control
```{r QC}
cat("Samples 2014\n")
# 2. Subset samples on missingness:
mat <- subset.samples.miss(data_old,Nmax=1, quiet=F)
# 3. Subset samples on zeros:
mat <- subset.samples.zero(mat,Nmax=1, quiet=F)
# 4. Subset samples on SD:
sds<-colSds(as.matrix(mat), na.rm = F)
names(sds)<-MET65
mat <- subset.samples.sd(as.matrix(mat),MEAN=colMeans(mat, na.rm=T),
                                SD=colSds(as.matrix(mat), na.rm=T),quiet=F,d=4)
  
orig_metabo65<-data.frame(mat)

cat("Samples 2020\n")
# 2. Subset samples on missingness:
mat <- subset.samples.miss(data_req,Nmax=1, quiet=F)
# 3. Subset samples on zeros:
mat <- subset.samples.zero(mat,Nmax=1, quiet=F)
# 4. Subset samples on SD:
sds<-colSds(as.matrix(mat), na.rm = T)
names(sds)<-MET65
mat <- subset.samples.sd(mat,MEAN=colMeans(mat, na.rm = T),
                                SD=sds,quiet=F, d=4)
req_metabo65<-data.frame(mat)

#Get the common samples
i<-intersect(rownames(orig_metabo65), rownames(req_metabo65))
data_old<-orig_metabo65[i,]
data_req<-req_metabo65[i,]

#Final dataset
cat("Final set\n")
dim(data_old)
```


# Correlations
For practicality we will visualize only the 65 metabolites that were selected to be the most uncorrelated and non derived.
```{r}
#If pyr is not present
MET65<-MET65[-which(MET65=="pyr")]

cor_65<-cor.assoc(data_old, data_req, MET65, method = "spearman")

cor_65<- cor_65[order(cor_65$cor),]
cor_65$met<-factor(cor_65$met, levels = cor_65$met)
ggplot(data=cor_65,aes(x=met,y=cor)) +
  geom_bar(stat="identity", fill="steelblue")+
  labs(title="Correlations old data and requantified",
       y="Correlations")+
  scale_y_continuous(breaks=seq(0, 1, by = 0.1))+
  theme(plot.title = element_text(size = 18, face = "bold",hjust = 0.5),
        axis.text.x=element_text(size=12,angle=45, hjust=1),
        axis.text.y=element_text(size=12),
        axis.title=element_text(size=16,face="bold"))
```

#Scatterplot Comparisons
```{r}
metcomparison<-foreach(i=MET65) %do%{
  axislimits<-c(min(c(data_old[,i],data_req[,i])), max(c(data_old[,i], data_req[,i])))
  
  pl<-density_scatterplot(x=data_old[,i],p=data_req[,i], xname = i, yname = paste("requantified",i), subtitle="", xlimits= axislimits, ylimits= axislimits)
}
names(metcomparison)<-MET65
```


# Mortality score
```{r}
mort_score <- comp.mort_score(data_old)

mort_score_req <- comp.mort_score(data_req)

cor(mort_score,mort_score_req, use = "pairwise.complete.obs")

mort_comparison<-density_scatterplot(x=mort_score,p=mort_score_req, xname = "data 2014",  yname = "data 2020", title="Comparison mortality score", subtitle="", xlimits= c(-4,4), ylimits= c(-4,4))

m_betas<-data.frame(mort_betas[,c(1,3)],cor=cor_65[mort_betas$Abbreviation,"cor"])
m_betas
```
